---
title: 'Week 8: Lab - Linear Modeling (Making Predictions)'
author: '[Grace Whitehurst]'
date: '[03/08/2026]'
output: html_document
---

---

# Instructions
The textbook’s chapter on linear models (“Line Up, Please”) introduces 
**linear predictive modeling** using the workhorse tool known as 
**multiple regression**. The term “multiple regression” has an odd history, 
dating back to an early scientific observation of a phenomenon called 
**“regression to the mean”**. These days, multiple regression is just an 
interesting name for using a simple linear modeling technique to measuring the 
connection between one or more **predictor variables** and an **outcome variable**. 

In this exercise, we are going to use an open dataset to explore antelope population.

This is the first exercise of the semester where there is no sample R code to 
help you along. Because you have had so much practice with R by now, you can 
create and/or find all of the code you need to accomplish these steps.

---

```{r setup, message = FALSE}
# Add your library below.

library(ggplot2)
library(readxl)

```


# Step 1 - Define "Model"
Write a definition of a model, based on how the author uses it in this chapter.

> [ For the chapter 16, the author said that in his textbook he uses model previously to define a repository. And then for this chapter, he says that he is now using it as defining mathematics for statistics specifically for prediction based analysis. "We want to discuss prediction models that are created from a statistical analysis process (in our case, from procedures offered by R). These models analyze data that the user supplies (such as a dataframe of observations) and then calculate a set of numerical coefficients that help us with prediction." (Saltz and Stanton, 197) (FOLLOW THE PRACTICE LINE-BY-LINE USING RSTUDIO!!!) ]


---


# Step 2 - Review the data 
You can find the data from Cengage's website. 
This URL will enable you to download the dataset into excel: 

* http://college.cengage.com/mathematics/brase/understandable_statistics/7e/students/datasets/mlr/excel/mlr01.xls

The more general website can be found at:

* http://college.cengage.com/mathematics/brase/understandable_statistics/7e/students/datasets/mlr/frames/frame.html

If you view this in a spreadsheet, you will find four columns of a small dataset:

* The first column shows the number of fawn in a given spring (fawn are baby antelope). 
* The second column shows the population of adult antelope.
* The third shows the annual precipitation that year. 
* And finally the last column shows how bad the winter was during that year.

```{r, "Step 2"}
# No code necessary; Just review the data.
```

---


# Step 3 - Read in the data
You have the option of saving the file to your computer and reading it into R, 
or you can read the data directly from the web into a dataframe.

```{r, "Step 3"}
# Write your code below.
myDF<-as.data.frame(readxl::read_excel("./data/mlr01.xls"))
colnames(myDF) <- c("nFawn", "adultPop", "precip", "winter")
head(myDF)
```


---


# Step 4 - Inspect the data
You should inspect the data using `str()` to make sure that 1) all the cases have
been read in (n=8 years of observations) and 2) that there are four variables.

```{r, "Step 4"}
# Write your code below.
str(myDF)
```


---


# Step 5 - Create bivariate plots
Create bivariate plots of the number of baby fawns versus adult antelope population,
precipitation that year, and severity of the winter.  
**Your code should produce three separate plots.** Make sure the y-axis and x-axis are labeled. 
Keeping in mind that the number of fawns is the outcome (or dependent) variable, 
which axis should it go on in your plots? You can also create scatter plots where 
size and colors reflect the two variables you didn't use 
(remember the visualization homework/lab. If you create these plots, you can earn extra 1 point).

Question: which variable is the most highly correlated with Fawn Count?

## Step 5.1 - Fawn Count by Adult Population
```{r, "Step 5.1"}
# Write your code below.
#install.packages("dplyr")
library(dplyr)
ggplot(myDF, aes(x=adultPop, y=nFawn)) +
       geom_point() +
       stat_smooth(method = "lm", col="white", fill="brown") +
       labs(title = "Fawn Count by Adult Population", 
         x = "Adult Population", 
         y = "Fawn Count") +
       theme_minimal()
```

## Step 5.2 - Fawn Count by Annual Precipitation
```{r, "Step 5.2"}
# Write your code below.
ggplot(myDF, aes(x=precip, y=nFawn)) +
       geom_point() +
       stat_smooth(method = "lm", col="tan", fill="lightgreen") +
       labs(title = "Fawn Count by Annual Precipitation", 
         x = "Annual Precipitation", 
         y = "Fawn Count") +
       theme_minimal()

```

## Step 5.3 - Fawn Count by Winter Severity Index
```{r, "Step 5.3"}
# Write your code below.
ggplot(myDF, aes(x=winter, y=nFawn)) +
       geom_point() +
       stat_smooth(method = "lm", col="tan", fill="lightblue") +
       labs(title = "Fawn Count by Winter Severity Index", 
         x = "Winter Severity Index", 
         y = "Fawn Count") +
       theme_minimal()

```


---


# Step 6 - Create regression models 
Create three regression models of increasing complexity using `lm()`, then
analyze the results. Based on the knowledge you've accumulated from Step 5, develop models.

* **Model one**: Fit the model to predict the number of fawns using the one variable you think is the most highly correlated with the outcome variable based on Step 5.
* **Model two**: Fit the model to predict the number of fawns using two variables (choose one variable to add to the one variable you selected in Model one)
* **Model three**: Fit the model to predict the number of fawns from all the three variables.  

## Step 6.1 - Predict Fawn Count using one input variable
```{r, "Step 6.1"}
# Write your code below.
first_M <- lm(formula = nFawn ~ precip, data = myDF)
summary(first_M)

```

## Step 6.2 - Predict Fawn Count using two input variable
```{r, "Step 6.2"}
# Write your code below.
second_M <- lm(formula = nFawn ~ precip + adultPop, data = myDF)
summary(second_M)

```

## Step 6.3 - Predict Fawn Count using three input variables
```{r, "Step 6.3"}
# Write your code below.
third_M <- lm(formula = nFawn ~ precip + adultPop + winter, data = myDF)
summary(third_M)

```

## Step 6.4 - Analysis
Which regression model works best? 
Which of the predictors are statistically significant in each model? 
If you wanted to create the most parsimonious model 
(i.e., the one that did the best job with the fewest predictors), 
what would it contain? ***You MUST answer these questions***.

> [>>>The regression model that worked the best was the third model that included all the variables with a R-squared value of 97%. This works the best because there is more information about the fawn count having correlation increase for each variable that together makes the fawn count supported in each category, with winter having the smallest influence. 
>>>>>>In the first model, the p-value was stastically significant as the value of 0.00104 meant strong significance. The second model, the p-value of precip at 0.235 as measuring adultPop as well measures fawn count as correlated closer to adultPop, but there was no significance. Lastly, in the third model, the R squared measures 97% significance of the model to Fawn Count increase. 
>>>>>>If I wanted to create the parsimonious model, I would choose the first model as it has the fewest predictors.

>>>>>>AI Disclosure Statement:
I did use AI Chatgpt this lesson to double check my work, because I had a feeling my original x and y axis labels were wrong. Here is my prompt: 
"on an x and y axis, if im measuring fawn count with each variable, which is x and which is y"
Chatgpt response: 
"Good question — this is a core regression concept, so get this clear.
Rule
Y-axis = the thing you are predicting (response variable)
X-axis = the variable doing the predicting (explanatory variable)
In your models
Your response variable is nFawn (number of fawns).
So:
Y-axis"
>>>>>>]


---


# Step 7 - Upload the compiled file
Please only include print outs of data sets using "head" function. 
I will take points off if you include more than two pages of dataset print outs.